### 2013-07-24 doc/asplos14/evaluation.tex
fixed some grappas

%\comment{
%\TODO{Grappa is well-suited to applications where there is
%    abundant data parallelism but unpredictable or imbalanced data
%    access. BSP is efficient where there is sufficient parallel slack.
%    Although Grappa loses some asynchrony by relying heavily on
%    aggregation, it still provides more flexibility for sections of the
%    computation to proceed more quickly. }
%
%\TODO{how much concurrency is needed on apps. Our basic plot of
%    performance vs workers (with and without aggro?)}
%\TODO{relationship between user-thread-visible latency and aggregate throughput.}
%\TODO{characterization of aggregation message size vs nodes}
%\TODO{show our scalability plots; if they don't scale, show
%    specifically why. It is okay if its because aggregation stops
%    working after a point if we can discuss future scalability.
%    Perhaps reference ``MPI on millions of cores'' type work}
%}
\end{itemize}

@@

%\TODO{go into more detail about what Grappa is doing when executing
%    GUPs. Specifically note what the rate per core is, what the
%    aggregation and deaggregation rates are. Show on the plot the
%    message size distribution (some box plot)}

@@

We compared this to a in-memory modification of the UPC implementation
of UTS \cite{UTS}, where each core hosts a UPC ``THREAD''. Berkeley
UPC (BUPC) is built upon GASNet, like Grappa, but it uses RDMA transfers directly
and the compiler overlaps remote communication when possible. We expect BUPC to
perform similarly on the T3 tree, where there is little parallelism to
overlap communication and computation, and no better than T3 on T1,
given the inability to overlap vertices on each core without
multithreading (a multithreaded variant of UPC has been evaluated in
\cite{HotSLAW}). Our experiments confirm that BUPC achieves the lower throughput of
Grappa on T3 for \emph{both} the T1 and T3 trees.

@@

In order to put \Grappa's performance into a general context, we compare it
with XMT running BFS, PageRank, IntSort, GUPS and UTS. Since XMT is a
different hardware platform, we also compare \Grappa with optimized MPI
versions of BFS and GUPS running on the same hardware. We run all experiments with 64 nodes and 16 cores per node.

@@

\begin{table}[htb]
\begin{center}
\begin{tabular}{l|c|c|c}
         & \Grappa & XMT   & MPI   \\ \hline
GUPS     & 1       & 2.23  & 0.11 \\ 
BFS      & 1       & 1.63  & 3.52 \\ 
IntSort  & 1       & 3.59  & 5.36 \\
UTS (T1) & 1       & 0.38  & --   \\  % 0.09 for UPC
Pagerank & 1       & 4.35  & 4.87 \\ 
\end{tabular}
\end{center}
\caption{Comparing \Grappa with XMT and optimized MPI. Numbers are presented as throughput on 64 nodes normalized to \Grappa.}
\label{tab:grappa-comparisons}
\end{table}

@@

\paragraph{UTS.}
It is perhaps not surprising that the XMT is faster than
\Grappa on many of the benchmarks; after all, the XMT is custom
hardware optimized for irregular applications. The XMT can
context-switch every cycle, while \Grappa takes $~$50 ns. The XMT can
complete 100 million random-access remote reads per second per node in
hardware, while \Grappa must spend many instructions managing and
aggregating each of these reads in software.

@@

Since \Grappa's tasks are implemented entirely in software, it is easy
to support arbitrary synchronization semantics. \Grappa allows both
XMT-like per-vertex task joins as well as whole-problem joins that are
a better fit for UTS. Furthermore, \Grappa's work-stealing distributed
task queue and recursive loop decomposition encourage locality in task
execution, so that the system is optimized for the common case of a
processor executing a task it spawned.

@@

\paragraph{BFS.}
The \Grappa and MPI BFS implementations are quite similar. They both
use the same graph representation, and they both depend on aggregation
for performance. The difference is that the BFS' aggregation is
integrated with the traversal code and specialized for the problem:
since the only messages required are edge traversals, the aggregation
buffers store only pairs of 8-byte vertex IDs.  In contrast, \Grappa
pays execution and space overhead in order to support aggregation of
arbitrary messages; the messages \Grappa aggregates include the two
8-byte vertex IDs along with 8 bytes of deserialization information as
well as 8 bytes of synchronization information. Furthermore, the MPI
implementation writes the edge information directly into the
aggregation buffer while \Grappa executes additional code to serialize
and deserialize the messages.

\paragraph{GUPS.}
The MPI GUPS implementation includes an implementation of aggregation,
but it is not optimized to work when the aggregated data exceeds
cache, and it has limited support for concurrent communication with
multiple destinations. \Grappa benefits from being optimized for both
of these cases.

\paragraph{Pagerank.}
The MPI Pagerank implementation is built on top of the
highly-optimized sparse matrix support in the Trilinos~\cite{trilinos}
library. In contrast, the \Grappa implementation is a straightforward nested
loop.

\paragraph{IntSort.}
Both MPI and \Grappa IntSort implement a bucket sort. For \Grappa,
distributing keys into buckets accounts for the bulk of the execution
time; \Grappa does this using asynchronous delegate operations to
write the keys directly to their final destination. The MPI
implementation, on the other hand, first does a local sort of the keys
on each core, and then uses an \texttt{MPI\_Alltoallv()} collective
operation to move all the keys to their destination nodes in
bulk. This is essentially specialized aggregation combined with
collective communication. When we measure the time spent in this
region of both benchmarks, we find that the the sort and
\texttt{MPI\_Alltoallv()} are $~5\times$ faster than the \Grappa
version using generic aggregation and point-to-point communication. An
interesting future direction would be to extend \Grappa to support
aggregation via collective operations.

\paragraph{Summary.}

@@

Figures~\ref{fig:grappa-scaling}, \ref{fig:grappa-gups}, and
\ref{fig:grappa-uts} show scaling out to 64 nodes. \Grappa scales best
on BFS and worst on IntSort, but is in general able to make use of
more nodes. Unfortunately memory limitations in our current network
library keep us from exploring scaling beyond 64 nodes; this will be
addressed in future work.

>>>>

%\comment{
%\TODO{\Grappa is well-suited to applications where there is
%    abundant data parallelism but unpredictable or imbalanced data
%    access. BSP is efficient where there is sufficient parallel slack.
%    Although \Grappa loses some asynchrony by relying heavily on
%    aggregation, it still provides more flexibility for sections of the
%    computation to proceed more quickly. }
%
%\TODO{how much concurrency is needed on apps. Our basic plot of
%    performance vs workers (with and without aggro?)}
%\TODO{relationship between user-thread-visible latency and aggregate throughput.}
%\TODO{characterization of aggregation message size vs nodes}
%\TODO{show our scalability plots; if they don't scale, show
%    specifically why. It is okay if its because aggregation stops
%    working after a point if we can discuss future scalability.
%    Perhaps reference ``MPI on millions of cores'' type work}
%}
\end{itemize}

@@

%\TODO{go into more detail about what \Grappa is doing when executing
%    GUPs. Specifically note what the rate per core is, what the
%    aggregation and deaggregation rates are. Show on the plot the
%    message size distribution (some box plot)}

@@

We compared this to a in-memory modification of the UPC implementation
of UTS \cite{UTS}, where each core hosts a UPC ``THREAD.'' Berkeley
UPC (BUPC) is built upon GASNet, like \Grappa, but it uses RDMA transfers directly
and the compiler overlaps remote communication when possible. We expect BUPC to
perform similarly on the T3 tree, where there is little parallelism to
overlap communication and computation, and no better than T3 on T1,
given the inability to overlap vertices on each core without
multithreading (a multithreaded variant of UPC has been evaluated in
\cite{HotSLAW}). Our experiments confirm that BUPC achieves the lower throughput of
\Grappa on T3 for \emph{both} the T1 and T3 trees.

@@

In order to put \Grappa's performance into a general context, we compare it
with XMT running BFS, PageRank, IntSort, GUPS, and UTS\@. Since XMT is a
different hardware platform, we also compare \Grappa with optimized MPI
versions of BFS and GUPS running on the same hardware. We run all experiments with 64 nodes and 16 cores per node.

@@

\begin{table}[htb]
\begin{center}
\begin{tabular}{l|c|c|c}
         & \Grappa & XMT   & MPI   \\ \hline
GUPS     & 1       & 2.23  & 0.11 \\ 
BFS      & 1       & 1.63  & 3.52 \\ 
IntSort  & 1       & 3.59  & 5.36 \\
UTS (T1) & 1       & 0.38  & --   \\  % 0.09 for UPC
Pagerank & 1       & 4.35  & 4.87 \\ 
\end{tabular}
\end{center}
\caption{Comparing \Grappa with XMT and optimized MPI\@. Numbers are presented as throughput on 64 nodes normalized to \Grappa.}
\label{tab:grappa-comparisons}
\end{table}

@@

\paragraph{UTS.}
It is perhaps not surprising that the XMT is faster than
\Grappa on many of the benchmarks; after all, the XMT is custom
hardware optimized for irregular applications. The XMT can
context-switch every cycle, while \Grappa takes $~$50ns. The XMT can
complete 100 million random-access remote reads per second per node in
hardware, while \Grappa must spend many instructions managing and
aggregating each of these reads in software.

@@

Since \Grappa's tasks are implemented entirely in software, it is easy
to support arbitrary synchronization semantics. \Grappa allows both
XMT-like per-vertex task joins as well as whole-problem joins that are
a better fit for UTS\@. Furthermore, \Grappa's work-stealing distributed
task queue and recursive loop decomposition encourage locality in task
execution, so that the system is optimized for the common case of a
processor executing a task it spawned.

@@

\paragraph{BFS}
The \Grappa and MPI BFS implementations are quite similar. They both
use the same graph representation, and they both depend on aggregation
for performance. The difference is that the BFS aggregation is
integrated with the traversal code and specialized for the problem:
since the only messages required are edge traversals, the aggregation
buffers store only pairs of 8-byte vertex IDs.  In contrast, \Grappa
pays execution and space overhead in order to support aggregation of
arbitrary messages; the messages \Grappa aggregates include the two
8-byte vertex IDs along with 8 bytes of deserialization information as
well as 8 bytes of synchronization information. Furthermore, the MPI
implementation writes the edge information directly into the
aggregation buffer while \Grappa executes additional code to serialize
and deserialize the messages.

\paragraph{GUPS}
The MPI GUPS implementation includes an implementation of aggregation,
but it is not optimized to work when the aggregated data exceeds
cache, and it has limited support for concurrent communication with
multiple destinations. \Grappa benefits from being optimized for both
of these cases.

\paragraph{Pagerank}
The MPI Pagerank implementation is built on top of the
highly-optimized sparse matrix support in the Trilinos~\cite{trilinos}
library. In contrast, the \Grappa implementation is a straightforward nested
loop.

\paragraph{IntSort}
Both MPI and \Grappa IntSort implement a bucket sort. For \Grappa,
distributing keys into buckets accounts for the bulk of the execution
time; \Grappa does this using asynchronous delegate operations to
write the keys directly to their final destination. The MPI
implementation, on the other hand, first does a local sort of the keys
on each core, and then uses an \texttt{MPI\_Alltoallv()} collective
operation to move all the keys to their destination nodes in
bulk. This is essentially specialized aggregation combined with
collective communication. When we measure the time spent in this
region of both benchmarks, we find that the the sort and
\texttt{MPI\_Alltoallv()} are $~5\times$ faster than the \Grappa
version using generic aggregation and point-to-point communication. An
interesting future direction would be to extend \Grappa to support
aggregation via collective operations.

\paragraph{Summary}

@@

Figures~\ref{fig:grappa-scaling}, \ref{fig:grappa-gups}, and
\ref{fig:grappa-uts} show scaling out to 64~nodes. \Grappa scales best
on BFS and worst on IntSort, but is in general able to make use of
more nodes. Unfortunately memory limitations in our current network
library keep us from exploring scaling beyond 64~nodes; this will be
addressed in future work.
